---
title: "Estimating local extrema using derivative Gaussian processes"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Estimating local extrema using derivative Gaussian processes}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this vignette, we provide worked examples on simulated data to demonstrates how to use the package `dgp` to compute the posterior density of local extrama using derivative Gaussian processes, which is discussed in [Li et al. (2023). Semiparametric Bayesian inference for local extrema of functions in the presence of noise](https://arxiv.org/abs/2103.10606).

```{r}
library(dgp)
```


## Simulated Data

All 100 simulated data sets for analysis in the paper are stored in `./data/sim_data_n100.RData`. The simulated data are generated and plotted using the script `./data-raw/gen_sim_data.R` as shown below.

```{r}
load("../data/sim_data_n100.RData", verbose = TRUE)
```

The object `YY` save the 100 simulated data sets. `x_a` and `x_b` are the end points of the domain of `x`. `cri_pts` stores the true stationary points, and `no_data` is the number of replicated data.


```{r}
source("../data-raw/seed.R")
source("../data-raw/gen_sim_data.R")
```

## Simulation study
The simulation study in the reference paper can be reproduced by running the script `dgp_theory.R`, including comparing the methods DGP Beta(1, 1), DGP Beta(2, 3), STS [(Kovac, 2007)](http://cran.nexr.com/web/packages/ftnonpar/index.html), and NKS (Song et. al, 2006).

The script `effect_n.R` analyzes the effect of sample size on the posterior distribution of the local extrema, and reproduces Figure 2 in the paper. The script `estimate_number_point.R` reproduces Figure 3.


```{r}
## required packages
# library(KernSmooth) ## for NKS
# library(ftnonpar) ## for STS ‘GNU Fortran (GCC) 8.2.0’
library(matrixcalc)
library(emulator)
```


First create objects needed. `H0_diff` is the distance matrix in the powered exponential kernel function. 
```{r}
H0_diff <- lapply(YY, function(d) {
    outer(as.vector(d$x), as.vector(d$x),
          FUN = function(x1, x2) (x1 - x2))
})

idx <- seq(x_a, x_b, length.out = 500)
x_test <- sort(seq(x_a, x_b, length.out = 100))
n_test <- length(x_test)
n <- length(YY[[1]]$x)
grid_t <- seq(x_a, x_b, length.out = 400)
type1err <- 0.05
```


Then estimate $\sigma$, and the hyperpameters $\tau$ and lengthscale $h$ in the kernel by maximizing the marginal log likelihood.
```{r}
library(parallel)
library(doParallel)
library(Rsolnp)
cl <- makeCluster(6, type = "FORK")
registerDoParallel(cl)
EB_gp <- foreach(k = 1:no_data) %dopar% {
    res <- Rsolnp::solnp(pars = c(1, 1, 1), fun = log_mar_lik_gp,
                         LB = c(0.0001, 0.0001, 0.0001),
                         UB = c(1 / 0.0001, 1 / 0.0001, 1 / 0.0001),
                         control = list(TOL = 1e-5, trace = 0),
                         y = YY[[k]]$y, H0 = H0_diff[[k]])
    res$par
}
stopCluster(cl)
EB_gp[[1]]
```



Given the hyperparameters, we apply the function `log_post_t_theory()` to compute the log posterior density of local extrema. The function requires the following arguments:

- `t`: The grid of $t$ for evaluating the density value.
- `y`: the response value vector
- `x`: the input value vector
- `Kff`: The covariance matrix formed by the kernel function
- `A`: The matrix $K_{ff} + n\lambda I$ needed for computing the density
- `lambda`: the parameter $\lambda = \sigma ^ 2 / (n\tau ^ 2)$
- `h`: the lengthscale parameter
- `sig`: the noise scale parameter
- `shape1`, `shape2`: the shape parameters of the beta prior on the stationary point.
- `a`, `b`: input domain.

The following is an example of the first simulated data set.

```{r}
#| echo: true
#| eval: true
## ============================================================================
## DGP beta(1, 1) by Li et al. (2023)
## ============================================================================
log_post <- rep(0, length(grid_t))
sig_gp <- EB_gp[[1]][1]
tau_gp <- EB_gp[[1]][2]
h_gp <- EB_gp[[1]][3]
lambda_gp <- sig_gp ^ 2 / (n * tau_gp ^ 2)
Kff_gp <- se_ker(H0 = H0_diff[[1]], tau = 1, h = h_gp)
A_gp <- Kff_gp + diag((n * lambda_gp), n)
for (i in 1:length(grid_t)) {
    log_post[i] <- 
        log_post_t_theory(t = grid_t[i], y = YY[[1]]$y, x = YY[[1]]$x,
                          Kff = Kff_gp, A = A_gp, lambda = lambda_gp,
                          h = h_gp, sig2 = sig_gp ^ 2, shape1 = 1, shape2 = 1,
                          a = x_a, b = x_b)
}
post_prob <- exp(log_post - max(log_post))
```



```{r}
par(mar = c(4, 4, 2, 1))
plot(grid_t, post_prob, type = "l", lwd = 2, main = "beta(1, 1) prior")
```

The density with $beta(2, 3)$ prior is shown below.

```{r}
#| echo: false
#| eval: true
## ============================================================================
## DGP beta(1, 1) by Li et al. (2023)
## ============================================================================
log_post <- rep(0, length(grid_t))
sig_gp <- EB_gp[[1]][1]
tau_gp <- EB_gp[[1]][2]
h_gp <- EB_gp[[1]][3]
lambda_gp <- sig_gp ^ 2 / (n * tau_gp ^ 2)
Kff_gp <- se_ker(H0 = H0_diff[[1]], tau = 1, h = h_gp)
A_gp <- Kff_gp + diag((n * lambda_gp), n)
for (i in 1:length(grid_t)) {
    log_post[i] <- 
        log_post_t_theory(t = grid_t[i], y = YY[[1]]$y, x = YY[[1]]$x,
                          Kff = Kff_gp, A = A_gp, lambda = lambda_gp,
                          h = h_gp, sig2 = sig_gp ^ 2, shape1 = 2, shape2 = 3,
                          a = x_a, b = x_b)
}
post_prob <- exp(log_post - max(log_post))
plot(grid_t, post_prob, type = "l", lwd = 2, main = "beta(2, 3) prior")
```


The function `get_hpd_interval_from_den()` obtains the $(1-\alpha)100\%$ highest posterior density (HPD) intervals from the density. The following calculate the HPD intervals for the first data set when $beta(2, 3)$ prior is used.


```{r}
#| echo: true
#| eval: true
hpdi <- get_hpd_interval_from_den(post_prob, grid_t = grid_t, target_prob = 0.95)
hpdi
cri_pts
```

There are three separated HPD intervals, each capturing one of the true critical points.


Another useful function is `get_map()` that obtains the maximum a posteriori (MAP) estimate from the posterior density.

```{r}
get_map(post_den = post_prob, grid_t = grid_t, hpdi = hpdi)
```

